---
title: "One big practice file: Day 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
library(lubridate)
library(broom)
```



# Read in files  

```{r}
# penguins
data(penguins)  # from the palmerpenguins package

# fish
fish_wq <- read.csv(here::here("data", "GNDfish_wq.csv"))
fish_sites <- read.csv(here::here("data", "GNDfish_sites.csv"))
fish_lengths <- read.csv(here::here("data", "GNDfish_lengths.csv"))

# bcwq
bc_jan <- read.csv(here::here("data", "GNDBCWQ012422.csv"),
                   skip = 8)
bc_feb <- read.csv(here::here("data", "GNDBCWQ022422.csv"),
                   skip = 8)

# clean up bcwq
bcwq <- bind_rows(bc_jan, bc_feb) 
names(bcwq) <- c("Date", "Time", "TimeSec",
                 "Site", "Depth", "DO_pct",
                 "DO_mgl", "pressure", "Sal",
                 "SpCond", "Turb", "Wiper",
                 "pH", "pH_mV", "Temp",
                 "Battery", "CablePwr")

bcwq <- bcwq %>% 
    mutate(DateTime = mdy_hms(paste(Date, Time)),   # this had to come before the date conversion itself
           Date = mdy(Date)) %>% 
    select(Date, DateTime, Depth, DO_pct, DO_mgl, Sal,
           SpCond, Turb, pH, Temp, Battery)
```



## On your data  

**Add your data files to the mix here too!** No better way to learn a new skill than to apply it to the thing **you've** been needing it for! And if it's not working on your dataset, this workshop is a great place to get help.  

```{r}


```



# Practice regression - and assessing assumptions   

## Ask a question  

Some questions you might answer:  

+  Is `body_mass_g` well-predicted by `flipper_length_mm` in Adelie penguins? If so, what is that relationship?    
+  What if you add in `bill_length_mm`, `bill_depth_mm`, and/or `sex`?  
+  Is there a relationship between `do_mgl` and `water_temp_c` in the `fish_wq` dataset?  
+  Does `water_temp_c` differ between seasons in the `fish_wq` dataset?  


How would you approach each of these questions? Before writing any code, write down and draw a picture of what you expect the relationship (if there is one) to look like.  

## EDA  

Make a couple of plots of the variables you plan to use, so you know what you're getting into.  

```{r}


```


## Code the model  

Do you want to use `lm()`? Would `aov()` work in your situation?  

```{r}


```


## Check assumptions  

+  **L**: linear relationship between your response and predictor(s)  
+  **I**: independent errors (residuals)  
+  **N**: normally distributed errors (residuals)  
+  **E**: equal variance of errors (residuals)  

```{r}


```


## Look at model output  

NOW think about p-values, slopes, and standard errors. Pull out the information you would report in a paper.  

```{r}


```

