---
title: "Base R Subsetting"
author: "Kim Cressman"
date: "5/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
```


# Subsetting Palmer Penguins  

First, let's explore the data frame.  

```{r}
glimpse(penguins)
summary(penguins)
```


## Choose columns based on position  

### Just one column  

```{r}
penguins[1]       # returns a tibble with one column
penguins[ , 1]    # returns a tibble with one column
penguins[[1]]     # returns the contents of the first column
```

### More than one column  

```{r}
penguins[1:2]          # returns a tibble with two columns
penguins[c(1, 2)]      # returns a tibble with two columns
   
penguins[ , 1:2]       # returns a tibble with two columns
penguins[ , c(1, 2)]   # returns a tibble with two columns
```



## Choose columns based on name  

### Just one  

```{r}
penguins["species"]      # returns a tibble with one column
penguins[ , "species"]   # returns a tibble with one column
penguins$species         # returns the contents of the column
```

### More than one  

```{r}
penguins[c("species", "island")]
penguins[ , c("species", "island")]
```

Can we use the `:` operator to denote a sequence?  

```{r, eval = FALSE}
penguins[c("species":"island")]
```

....not in base R.  


We can use `which()` to find the positions of these columns, however. (You can see why you might not want to do this, and you probably won't do it frequently. But knowing that there's a way could come in handy for you someday.)  

```{r}
which(names(penguins) == "species")
which(names(penguins) == "island")

penguins[which(names(penguins) == "species"):which(names(penguins) == "island")]
```


## Choose rows based on position  

```{r}
penguins[1, ]
penguins[3, ]
```


## Choose row/column combinations, based on position  

[row, column]

```{r}
penguins[1, 3]   # first row, 3rd column
penguins[3, 1]   # third row, 1st column  

penguins[1:3, 1:3]  # first three of each dimension
```


## Choose rows based on a condition  

Specify a condition that will be TRUE or FALSE, and rows that satisfy the condition (TRUE) will be returned.  

```{r}
penguins[penguins$species == "Adelie", ]
penguins[penguins$species != "Adelie", ]

penguins[penguins$island == "Torgersen", ]
penguins[penguins$bill_length_mm < 40, ]
```



### by position  

### by conditions  

Say we only want the bill length of Adelie penguins.  

```{r}
penguins[penguins$species == "Adelie", "bill_length_mm"]

```

