---
title: "Statistics: Running and exploring outputs"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
library(lubridate)
library(broom)
```



# Read in files  

```{r}
# penguins
data(penguins)  # from the palmerpenguins package

# bcwq
bc_jan <- read.csv(here::here("data", "GNDBCWQ012422.csv"),
                   skip = 8)
bc_feb <- read.csv(here::here("data", "GNDBCWQ022422.csv"),
                   skip = 8)

# fish
fish_wq <- read.csv(here::here("data", "GNDfish_wq.csv"))
```


## On your data  

**Add your data files to the mix here too!** No better way to learn a new skill than to apply it to the thing **you've** been needing it for! And if it's not working on your dataset, this workshop is a great place to get help.  

```{r}


```



# An example of Simple Linear Regression  

In which we know some things: DO mg/L is a calculated value, based on DO % saturation, temperature, and salinity.  

## Ask a question  

Our research questions here may be:  
+  is there a linear relationship between DO mg/L and DO %?  
+  what is the slope of that relationship?  

We expect this to be a pretty straight line, but not perfect (because salinity and temperature matter). Let's see what this looks like, in various ways.  

## Clean up bcwq data  

```{r}
bcwq <- bind_rows(bc_jan, bc_feb) 
names(bcwq) <- c("Date", "Time", "TimeSec",
                 "Site", "Depth", "DO_pct",
                 "DO_mgl", "pressure", "Sal",
                 "SpCond", "Turb", "Wiper",
                 "pH", "pH_mV", "Temp",
                 "Battery", "CablePwr")

bcwq <- bcwq %>% 
    mutate(DateTime = mdy_hms(paste(Date, Time)),   # this had to come before the date converstion itself
           Date = mdy(Date)) %>% 
    select(Date, DateTime, Depth, DO_pct, DO_mgl, Sal,
           SpCond, Turb, pH, Temp, Battery)
```

## EDA  

Know what we're getting into. See if there are outliers, and determine what to do about them (note: just deleting them, for the sole fact that they're outliers, is NOT appropriate - you need to understand what caused them).  

```{r}
plot(bcwq$DO_pct)
plot(bcwq$DO_mgl)
plot(bcwq$pH)
plot(bcwq$Sal)
```

Ah, the sonde came out of water a couple times - this happens on very low tides at this station. **Reminder that knowing your data is important!** As is exploring it this way.  

For simplicity, we'll exclude rows where DO_mgl is greater than 12. In the real world, you would want to be more careful about verifying and excluding potential out-of-water points (for SWMP, we do it through our QA/QC flag and code columns).  


```{r}
bcwq_qaqc <- bcwq %>% 
    filter(DO_mgl < 12)
plot(bcwq_qaqc$DO_mgl)
```



## Run the regression  


One important assumption of linear regression is that **the relationship in question is actually linear** - not curved. You should be able to draw a straight line through your points.  


```{r}
fit1 <- lm(DO_mgl ~ DO_pct, data = bcwq_qaqc)
plot(DO_mgl ~ DO_pct, data = bcwq_qaqc)
```


## Check assumptions  

Penn State's STAT 501 course notes have very good information on assessing assumptions of linear prediction: https://online.stat.psu.edu/stat501/lesson/4  

They actually lay out all the assumptions above using the acronym **LINE**:  

+  **L**: linear relationship between your response and predictor(s)  
+  **I**: independent errors (residuals) (no patterns like "bigger residuals when predictions are bigger"; no autocorrelation)  
+  **N**: normally distributed errors (residuals)  
+  **E**: equal variance of errors (residuals) 


We assessed the **L** condition above, with a plot of our data. Now to check the other conditions by looking at plots of residuals:  


```{r}
hist(residuals(fit1), breaks = 20)

# we don't want to see patterns in residuals: the next couple of plots 
# should just be random scatter.
# We *do* want the points in the QQ plot to follow a line.
plot(fit1)
```


Okay, these assumptions look reasonable.  


### Something that (should) look bad:  

I'm going to make up some data that has a relationship, but not a simple linear one. This one is curved. We'll investigate the residuals of this too, so you can see what bad ones look like.    

```{r}
x <- seq(from = 1, to = 50, length.out = 1000)
y <- 2*(x^2) + x - rnorm(n = length(x), mean = 10, sd = 100)
plot(y ~ x)
fit2 <- lm(y ~ x)
hist(residuals(fit2), breaks = 20)
plot(fit2)
```


**Notice: we haven't even mentioned p-values yet.**  

**It is critical to assess assumptions of your statistical model to make sure it is appropriate BEFORE drawing any inferences.**  

If assumptions aren't met (especially the normality one), there are other statistical methods you can use! Many universities have a statistical consulting center; reach out!  

Our model above was appropriate though, so we'll continue.  

## Investigate model output  

Now it's time to look at output:  

```{r}
fit1
```

All this means is that, if we were to write the equation describing the two parameters, it would look like:  

DO_pct = 0.089*DO_mgL - 0.126

We get lots more information from the `summary()` function:  

```{r}
summary(fit1)
fit1_summ <- summary(fit1)
```

And if you have categorical predictors (is there a difference in size between group A and group B?) and just want to run an ANOVA, you get the ANOVA table thusly:  

```{r}
anova(fit1)
```


### Remember our friend, the list?  

You probably don't want to copy and paste, or re-type, all the coefficients and p-values you get. It is useful to be able to pull out the pieces of output you want. Statistical output is stored in a list format; and exactly how it is structured depends on the package and the test that you are using.  

```{r}
# coefficients
fit1$coefficients

# just want the slope?  
fit1$coefficients[2]

# p-values etc. from the summary:

# more detail from the coefficients table: standard errors etc.
fit1_summ$coefficients

fit1_summ$r.squared

fit1_summ$adj.r.squared
```


### The easier way: {broom}  

```{r}
broom::glance(fit1)
broom::tidy(fit1)
```


This gets even more useful with multiple linear regression, or more complicated models.  


# ANOVA (categorical predictors)  

You can run ANOVA using the `lm()` format above, followed by `anova()` on the output. There's a bit of a shortcut: `aov()`. This is what's known as a wrapper function for the above.  


## Ask a question  

Research question: is body weight different between male and female Gentoo penguins?  

```{r}
gent <- penguins %>% 
    filter(species == "Gentoo")
```

## EDA  

```{r}
boxplot(body_mass_g ~ sex, gent)
```

## Code the model  

```{r}
fit3 <- aov(body_mass_g ~ sex, gent)
```

## Check assumptions  

```{r}
# don't forget to check residuals:
plot(fit3)
```

## Look at model output  

```{r}

fit3
summary(fit3)

glance(fit3)
tidy(fit3)

# only real problem with {broom} is it doesn't tell you what kind of statistic you're using
# from summary(fit3), we see that it's F; with 1 and 117 degrees of freedom
```


# Don't forget: care about more than just a p-value  

Other things to report are: sample size (from `glance`: 119, or get this from your data); the test statistic (from `summary`: F = 216.7 on 1 and 117 degrees of freedom); R^2 (0.649).  

What are the mean and standard deviations of weights for both sexes? We found a statistically significant difference; how big is the difference in reality?   

```{r}
gent %>% 
    filter(!is.na(body_mass_g),
           !is.na(sex)) %>% 
    group_by(sex) %>% 
    summarize(mean_wt = mean(body_mass_g),
              sd_wt = sd(body_mass_g),
              min_wt = min(body_mass_g),
              max_wt = max(body_mass_g))
```


# A word on factors  

You need to use 'factors' if you have categories that look like numbers. You don't want a line with a slope in the following scenario: you want to know if there are differences in categories.  

```{r}
plot(salinity_ppt ~ site, fish_wq)

# if we don't use factors:  
fit5 <- lm(salinity_ppt ~ site, fish_wq)
summary(fit5)
# we get an overall slope and intercept

plot(fit5)


fit6 <- lm(salinity_ppt ~ as.factor(site), fish_wq)
summary(fit6)
# different estimates for each site

anova(fit6)

# alternately:
summary(aov(salinity_ppt ~ as.factor(site), fish_wq))
```