---
title: "WQ data"
author: "Kim Cressman"
date: "5/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
```

# Read in data  

## knowing how many extra header rows there are  

```{r}
bc_jan <- read.csv(here::here("data", "GNDBCWQ012422.csv"),
                   skip = 8)
```

## NOT knowing how many rows there are, but knowing a value in the row you want to be the header  

What is in our desired header row, that doesn't show up before? `Date`! Let's look for that in the first column.  

Remember subsetting with square brackets? This is a situation where that's potentially easier than using `dplyr::select()`.  

```{r}
bc_test <- read.csv(here::here("data", "GNDBCWQ012422.csv"))
str_which(bc_test[ , 1], "Date") # from the stringr package. grep is comparable in base R.
```

Now we've found it. Notice that `read.csv()` uses the first row of your file as headers by default, and we didn't turn that off: so row 8 in our data frame is actually row 9 of our .csv file, and we want to skip 8 rows to make that one the header.  

I like to assign whatever I've found to a variable, and use *that* in the `skip` argument. You could use the `str_which` code there, but it gets messy. And more complicated if you later decide you need to figure out a different string to search for.  

```{r}
header_index <- str_which(bc_test[ , 1], "Date")

bc_test2 <- read.csv(here::here("data", "GNDBCWQ012422.csv"),
                   skip = header_index)

# once you're satisfied that bc_test2 matches bc_jan, remove the test data frames
# rm(bc_test, bc_test2)
```


## Reading in the 2nd file and binding them  

Not sure if it's the same number of header rows, so I'm going to do the header_index thing again.

```{r}
test <- read.csv(here::here("data", "GNDBCWQ022422.csv"))
header_index <- str_which(test[ , 1], "Date")
rm(test)

bc_feb <- read.csv(here::here("data", "GNDBCWQ022422.csv"),
                   skip = header_index)
```

The names look gross but they're consistent, so we'll bind the files together before dealing with names.  

```{r}
bc_all <- bind_rows(bc_jan, bc_feb)
```


Deal with names. We could select and rename at the same time with `dplyr::select()`:  

```{r}
# first figure out what we're dealing with
names(bc_all)

bc_all %>% select(Site = Site.Name,
                  Date = Date..MM.DD.YYYY.) %>% 
    head()
```

I do that a LOT! But in this case, doing so requires us to type out all the old, difficult names. So for this data, I would select by position - using base R! - and then rename.  

```{r}
names(bc_all)

bc_all2 <- bc_all[ , c(1:2, 5:7, 9:11, 13, 15:16)]

# make sure I got the right columns; counting is hard
glimpse(bc_all2)    

# and rename:
names(bc_all2) <- c("Date",
                    "Time",
                    "Depth_m",
                    "DO_pct",
                    "DO_mgl",
                    "Sal_psu",
                    "SpCond",
                    "Turb",
                    "ph",
                    "Temp_C",
                    "Battery_V")

glimpse(bc_all2)
```

Oops! I typed pH as lower case! It's also possible to change just one name, using good old square bracket indexing:  

```{r}
names(bc_all2)[9] <- "pH"
names(bc_all2)
```

That involved counting though; if you know the name you're trying to change and it's easy to type, you can use `dplyr::rename()`:  

```{r}
# first make my 'oops' names again:
names(bc_all2) <- c("Date",
                    "Time",
                    "Depth_m",
                    "DO_pct",
                    "DO_mgl",
                    "Sal_psu",
                    "SpCond",
                    "Turb",
                    "ph",
                    "Temp_C",
                    "Battery_V")
# confirm
names(bc_all2)

# rename using dplyr
bc_all2 <- bc_all2 %>% 
    rename(pH = ph)
names(bc_all2)
```

That involved more typing, this time. But it can be useful when you're trying to make a lot of changes to your data frame at the same time.  


# Group and summarize  

Say we want daily average salinity:  

```{r}
bc_daily_sal <- bc_all2 %>% 
    group_by(Date) %>% 
    summarize(mean_Sal = mean(Sal_psu, na.rm = TRUE))
```

Say we want daily average salinity and DO_mgl:  

```{r}
bc_daily2 <- bc_all2 %>% 
    group_by(Date) %>% 
    summarize(mean_Sal = mean(Sal_psu, na.rm = TRUE),
              mean_DO = mean(DO_mgl, na.rm = TRUE))
```

Let's see what this looks like.  

```{r}
ggplot(bc_daily2) +
    geom_line(aes(x = Date, y = mean_Sal))
```

That got weird, huh?  




Did you notice every time we ran `glimpse()` on this data frame, Date was a character? We need to turn it into an actual date. I like to do this as I'm reading a file in - verify that all the columns are the way I want them to be - but I wanted you to see what happens when you forget. If you get a graph like this, you should think to yourself, "Date is wrong."  

Let's actually fix this in `bc_all2`, then re-make `bc_daily2`.  

```{r}
bc_all2 <- bc_all2 %>% 
    mutate(Date = mdy(Date))
glimpse(bc_all2)

bc_daily2 <- bc_all2 %>% 
    group_by(Date) %>% 
    summarize(mean_Sal = mean(Sal_psu, na.rm = TRUE),
              mean_DO = mean(DO_mgl, na.rm = TRUE))
```

Now let's try this again.  

```{r}
ggplot(bc_daily2) +
    geom_line(aes(x = Date, y = mean_Sal, color = "Sal")) +
    geom_line(aes(x = Date, y = mean_DO, color = "DO"))
```


### Summarize ALL THE THINGS  

```{r}
bc_daily_all <- bc_all2 %>% 
    group_by(Date) %>% 
    summarize(across(Depth_m:Battery_V,  # : makes a sequence
                     mean, na.rm = TRUE))
```

With a little more effort, and some reading of the help file, we can get multiple summaries for each variable.  

```{r}
# ?across

bc_daily_all <- bc_all2 %>% 
    group_by(Date) %>% 
    summarize(across(Depth_m:Battery_V,
                     .fns = list(mean = ~mean(., na.rm = TRUE),
                                 min = ~min(., na.rm = TRUE),
                                 max = ~max(., na.rm = TRUE)
                                 )
                     )
              )
```

This can lead to some cool graphs:  

```{r}
ggplot(bc_daily_all) +
    geom_line(aes(x = Date, y = Sal_psu_min), col = "gray30") +
    geom_line(aes(x = Date, y = Sal_psu_max), col = "gray30") +
    geom_line(aes(x = Date, y = Sal_psu_mean), col = "blue", size = 2) +
    labs(title = "Daily average, max, and min salinity at BC in early 2022",
         y = "Salinity (psu)") +
    theme_bw()
```

